Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment

### 2017-04-18 Latex/article.tex
Virtually done.

% \usepackage[firstpage]{draftwatermark}
% \SetWatermarkFontSize{40pt}
% \SetWatermarkHorCenter{300pt}
% \SetWatermarkColor[rgb]{1,0.2,0.2}
% \SetWatermarkText{\textsf{Submitted April 2017.}}

@@

  \author*[1]{Roland Schäfer}
  %\runningauthor{...}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{In this paper, a case alternation occurring in German measure noun phrases is examined from the perspective of cognitively oriented corpus linguistics.
    In pseudo-partitive constructions such as \textit{eine Tasse guter Kaffe} `a cup of good coffee', the embedded kind-denoting noun either agrees in its case with the measure head noun or it is a genitive as in \textit{eine Tasse guten Kaffees}.
    This alternation occurs only under highly specific syntactic conditions.
    The analysis is formulated in terms of Prototype Theory. 
    I assume that the alternants are modelled after non-alternating neighbouring constructions, which represent the prototypes more purely.
    Primarily, it is argued that the frequencies with which individual lemmas occur in the prototypical constructions partially predict which alternant is chosen in the alternating case.
    Another major prediction is that the genitive construction is prototypical of more strongly grammaticalised measure nouns.
    I present a large-scale corpus study using the DECOW web corpus which supports the theoretical analysis.
    In the statistical analysis of the corpus study, it is discussed whether Bayesian estimators lead to substantially different results compared to Likelihood Maximisation methods in studies like this one.
    The model is estimated with Maximum Likelihood and Bayesian methods for comparison.
    Finally, a forced choice experiment and a self-paced reading experiment are reported.
    The usage-based findings from the corpus study show a clear correlation with the behaviour of cognitive agents in both experiments.
    The present study therefore contributes to a now well-established area of research into alternations using both corpus and experimental methods.}
  \keywords{corpus methods and experimental methods, self-paced reading, forced-choice, alternations, hierarchical models, measure constructions, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{Submitted April 2017. Do not quote.}
  \journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

For such probabilistic generalisations estimated on corpus data, there has always been an interest in correlating the findings with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data'', citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is'' (\textit{generally}) \textit{realized by native speakers}.
This position, which I adopt, does not, however, imply that we can in some way \textit{deduce mental representations from patterns of use}, \ie\ from corpus data, as  \citet[486--487]{Dabrowska2016} convincingly argues.%
\footnote{In its colloquial meaning, the verb \textit{deduce} seems appropriate here.
However, in terms of the scientific method, we are most likely talking about \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible, and it is the same for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
Technical problems of statistical inference aside (see also Section~\ref{sec:rightstatistics}), a successful experiment corroborates the theory, and a failed experiment weakens the theory.
The simplest assumption would be that the same approach is adequate for corpus studies.

The present study is conducted within the general paradigm of cognitive corpus linguistics and includes a comparison of the corpus findings with results from two experiments.
As a theoretical framework, I assume a model of similarity-based classification in the form of Prototype Theory.
Protoype Theory has been used in alternation research in cognitively oriented corpus linguistics (see \citealp{DivjakArppe2013}, \citealp{Gries2003}).
It is assumed that the variables annotated in a corpus study on an alternation phenomenon define prototypes for the alternants.
The similarity of a given feature vector (in a concrete sentence where one of the alternants is used) influences to a large extent which alternant is chosen by speakers.
This entails that a variant of Prototype Theory with features is assumed \citep{Rosch1978}, instead of the monolithic prototypes of earlier versions.
Prototype Theory is well suited for modeling constructional choices, but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types off effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation (Prototype Theory) or try to do without them (Exemplar Theory).
I am sceptical that corpus analysis alone could ever decide which theory is more adequate, given that substantial doubt has been voiced whether even experimental methods are ultimately able to do so \citep{Barsalou1990}.%
\footnote{In \cite{DivjakArppe2013}, it was shown using corpus data how both models form a converging picture.}
Prototype Theory (with feature abstractions) is preferred here simply because it fits the established alternation modeling paradigm, which relies on features being weighted in statistical models.
In Section~\ref{sec:analyses}, a prototype-theoretical parlance is therefore adopted.

In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then, the corpus study is presented, including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments that correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun that specifies a quantity of the objects or the substance denoted by the kind-denoting noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
For illustration purposes, in the English \textit{a glass of good wine}, \textit{glass} is the measure noun, and \textit{wine} is the kind noun.
Measure nouns can be all sorts of nouns which denote a quantity (such as \textit{litre} or \textit{amount}) but also those denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \citet[284]{Brems2003} considers such nouns denoting containers etc.\ as measure nouns ``which, strictly speaking, do not designate a `measure', but display a more nebulous potential for quantification'' (see also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).

@@

@@

The alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the PGC in (\ref{ex:intro:alternation1}) and the NAC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are by-and-large functionally and semantically equivalent.
However, Section~\ref{sec:analyses} is devoted to developing hypotheses about subtle differences between them.%
\footnote{Some descriptive and normative grammars take stronger positions w.\,r.\,t.\ the acceptability of the two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below (especially Section~\ref{sec:exp:fc}) should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be unacceptable.
}

@@

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.
In principle, this paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.
However, in Section~\ref{sec:analyses}, the outer columns (\NACb\ and \PGCd) will still play a major role when the factors controlling the alternation are named.

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the \NACa\ and the \PGCa,}
%  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives,}
%  \item{\label{it:intro:femsg} case syncretism in feminine NPs,}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns,}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns,}
  \item{\label{it:intro:other} alternative constructions for expressing quantities.}
\end{enumerate}

@@

It was shown empirically in \cite{Hentschel1993} that such sentences are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it virtually impossible to study them using corpus linguistic methods (Section~\ref{sec:corpusstudies}), and I consequently ignore them.

@@

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (794 of 861 cases, or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular.%
\footnote{More concretely, if these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which have no normal plural forms at all.%
\footnote{Plurals \textit{Pfunde} and \textit{Prozente} have special meanings and have very restricted uses, mostly in idiomatic expressions and prefabs.
They cannot be used in MNPs.}
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected version \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected version \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected version denotes the quantity of wine fitting into two glasses, the inflected version denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, however, degrees of grammaticalisation as a factor influencing the alternation will be discussed.

@@

These construction have very idiosyncratic properties.
The construction with \textit{voller} is discussed in \cite{Zeldes2018} and the construction with \textit{mit} is discussed in \cite{Bhatt1990}.
They are not semantically equivalent to the NAC and the PGC, and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers, and the whole NP always refers to the container, and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts, and they are thus decidedly \textit{not} more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined alternants.

This concludes the descriptive overview of the phenomenon.
I have demonstrated that there is an alternation between two measure noun constructions in a narrow syntactic configuration (kind NP with an adjective but without a determiner), and that the two constructions differ in the case of the kind noun (case agreement with the measure noun or genitive).
I turn to some more theory-oriented discussion in the next section.

@@

First, my analysis is based on the idea that in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, %the first question is which noun constitutes the head of the whole construction.
%This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
%Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
%Furthermore, the internal structure of MNPs can only be understood if the strong\slash weak inflection patterns of adjectives are taken into account.
the strong\slash weak inflection patterns of adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), on the other hand, attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.

\begin{exe}
  \ex\label{ex:starkschwach} 
  \begin{xlist}
    \ex{\label{ex:starkschwach:a}\gll mit ein\textbf{-em} stark\textbf{-en} Kaffe\\
                                      with a strong coffee\\}
    \ex{\label{ex:starkschwach:b}\gll mit stark\textbf{-em} Kaffe\\
                                      with strong coffee\\}
  \end{xlist}
\end{exe}

@@

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collostructional Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least partly from the relative frequency with which the same lemmas are used in the non-alternating constructions.
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be used in the corpus study Section~\ref{sec:corpusstudies}.

@@

Third, as described above, the grammaticalisation path leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- inasmuch as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified.
However, it is much more prototypical of quantities to be specified with numerical precision.
This is most obvious with the highly grammaticalised physical measure nouns like \textit{centilitre} which are very prototypically used with exact numeral instead of unspecific quantifiers although both options are available (\textit{three centilitres} vs.\ \textit{several centilitres}).
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected alternants under this hypothesis.

@@

Fourth and finally, it was found that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising inasmuch as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using two very simple proxy variables.

@@

% In \citet[10]{GelmanShalizi2013}, the authors -- prominent Bayesians themselves -- acknowledge that a theory of statistical testing is a desideratum, and they state about the standard inductive interpretation of Bayesianism that ``most of this received view of Bayesian inference is wrong'', and they develop a Bayesian notion of p values (see also \citealp{Mayo2013}, for a frequentist reply; also \citealp{Senn2011} on different strands of Bayesianism and their stance on inductive vs.\ deductive reasoning, and \citealp{Mayo2011}, for a critical reply to \citealp{Senn2011}).

>>>>

\usepackage[firstpage]{draftwatermark}
\SetWatermarkFontSize{40pt}
\SetWatermarkHorCenter{300pt}
\SetWatermarkColor[rgb]{1,0.2,0.2}
\SetWatermarkText{\textsf{Submitted April 2017.}}

@@

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer. Submitted April 2017. Do not quote.}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{In this paper, a case alternation occurring in German measure noun phrases is examined from the perspective of cognitively oriented corpus linguistics.
    In pseudo-partitive constructions such as \textit{eine Tasse guter Kaffee} `a cup of good coffee', the embedded kind-denoting noun either agrees in its case with the measure head noun or it stands in the genitive as in \textit{eine Tasse guten Kaffees} with no difference in meaning.
    This alternation occurs only under specific syntactic conditions.
    The analysis is formulated in terms of Prototype Theory. 
    I assume that the alternants are modelled after non-alternating neighbouring constructions, which represent the prototypes more directly.
    Primarily, it is argued that the frequencies with which individual lemmas occur in the prototypical constructions partially predict which alternant is chosen in the alternating case.
    Another major assumption is that the genitive construction is prototypical of more strongly grammaticalised measure nouns.
    I present a large-scale corpus study using the DECOW web corpus which supports the theory.
    In the statistical analysis of the corpus study, it is discussed whether Bayesian estimators lead to substantially different results than Likelihood Maximisation methods in this and similar studies.
    The model parameters are estimated with both methods for comparison.
    Finally, a forced choice experiment and a self-paced reading experiment are reported.
    The usage-based findings from the corpus study show a correlation with the behaviour of cognitive agents in both experiments.
    The present study therefore contributes to a now well-established area of research into alternations using both corpus and experimental methods.}
  \keywords{corpus methods and experimental methods, self-paced reading, forced-choice, alternations, hierarchical models, measure constructions, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{Submitted April 2017. Do not quote.}
  \journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

For such probabilistic generalisations estimated on corpus data, there has always been an interest in correlating the findings with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.%
\footnote{In its colloquial meaning, the verb \textit{deduce} seems appropriate here.
However, in terms of the scientific method, we are most likely talking about \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
Technical problems of statistical inference aside (see also Section~\ref{sec:rightstatistics}), a successful experiment corroborates the theory, and a failed experiment weakens the theory.
The simplest assumption would be that the same approach is adequate for corpus studies.

The present study is conducted within the general paradigm of cognitive corpus linguistics and includes a comparison of the corpus findings with results from two experiments.
As a theoretical framework, I assume a model of similarity-based classification in the form of Prototype Theory.
Protoype Theory has been used in alternation research in cognitively oriented corpus linguistics (see \citealp{DivjakArppe2013}, \citealp{Gries2003}).
It is assumed that the variables annotated in a corpus study on an alternation phenomenon define prototypes for the alternants.
The similarity of a given feature vector (in a concrete sentence where one of the alternants is used) influences to a large extent which alternant is chosen by speakers.
This entails that a variant of Prototype Theory with features is assumed \citep{Rosch1978}, instead of the monolithic prototypes of earlier versions.
Prototype Theory is well suited for modeling constructional choices but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types off effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation (Prototype Theory) or try to do without them (Exemplar Theory).
I am sceptical that corpus analysis alone could ever decide which theory is more suitable, given that substantial doubt has been voiced whether even experimental methods are ultimately able to do so \citep{Barsalou1990}.%
\footnote{In \cite{DivjakArppe2013}, it was shown using corpus data how both models form a converging picture.}
Prototype Theory (with feature abstractions) is preferred here simply because it fits the established alternation modeling paradigm, which relies on features being weighted in statistical models.
In Section~\ref{sec:analyses}, a prototype-theoretical parlance is therefore adopted.

In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then the corpus study is presented, including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments that correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun that specifies a quantity of the objects or the substance denoted by the kind-denoting noun.
I call the kind-denoting noun the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
For illustration purposes, in the English \textit{a glass of good wine}, \textit{glass} is the measure noun and \textit{wine} is the kind noun.
Measure nouns can be all sorts of nouns which denote a quantity (such as \textit{litre} or \textit{amount}) but also those denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
Like \citet[284]{Brems2003}, I consider nouns as measure nouns ``which, strictly speaking, do not designate a `measure', but display a more nebulous potential for quantification'' (also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).

@@

\enlargethispage{1\baselineskip}

@@

The alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the PGC in (\ref{ex:intro:alternation1}) and the NAC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are by-and-large functionally and semantically equivalent.
However, Section~\ref{sec:analyses} is devoted to developing hypotheses about subtle differences between them.%
\footnote{Some descriptive and normative grammars take stronger positions with regard to the acceptability of the two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below (especially Section~\ref{sec:exp:fc}) should render it clear, there might be preferences, but we cannot assume either construction to be unacceptable.
}

@@

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP is summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb\ and the partitive genitive with a determiner in the kind NP the \PGCd.
For the alternants with an adjective but no determiner in the kind noun phrase I use the terms \NACa and \PGCa.
In principle, this paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.
However, in Section~\ref{sec:analyses}, the outer columns (\NACb\ and \PGCd) will still play a major role when the factors controlling the alternation are named.

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the \NACa\ and the \PGCa,}
%  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives,}
%  \item{\label{it:intro:femsg} case syncretism in feminine NPs,}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns,}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns, and}
  \item{\label{it:intro:other} alternative constructions for expressing quantities.}
\end{enumerate}

@@

It was shown empirically in \cite{Hentschel1993} that such sentences are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it virtually impossible to study them using corpus linguistic methods (Section~\ref{sec:corpusstudies}) and I consequently ignore them.

@@

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (794 of 861 cases, or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular.
%\footnote{More concretely, if these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear in their non-inflected form.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which have no normal plural forms at all.%
\footnote{Plurals \textit{Pfunde} and \textit{Prozente} have special meanings and have very restricted uses, mostly in idiomatic expressions and prefabs.
They cannot be used in MNPs.}
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected version \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected version \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected version denotes the quantity of wine fitting into two glasses, the inflected version denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, however, degrees of grammaticalisation as a factor influencing the alternation will be discussed.

@@

These construction have very idiosyncratic properties.
The construction with \textit{voller} is discussed in \cite{Zeldes2018} and the construction with \textit{mit} is discussed in \cite{Bhatt1990}.
They are not semantically equivalent to the NAC and the PGC and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers (where the whole NP always refers to the container) and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts and they are thus decidedly \textit{not} more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined alternants.

This concludes the descriptive overview of the phenomenon.
I have demonstrated that there is an alternation between two measure noun constructions in a narrow syntactic configuration (kind NP with an adjective but without a determiner), and that the two constructions differ in the case of the kind noun (case agreement with the measure noun or genitive).
I turn to more theory-oriented discussion in the next section.

@@

First, my analysis is based on the idea that in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, %the first question is which noun constitutes the head of the whole construction.
%This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
%Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
%Furthermore, the internal structure of MNPs can only be understood if the strong\slash weak inflection patterns of adjectives are taken into account.
the strong\slash weak inflection patterns of adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.

\begin{exe}
  \ex\label{ex:starkschwach} 
  \begin{xlist}
    \ex{\label{ex:starkschwach:a}\gll mit ein\textbf{-em} stark\textbf{-en} Kaffee\\
                                      with a strong coffee\\}
    \ex{\label{ex:starkschwach:b}\gll mit stark\textbf{-em} Kaffee\\
                                      with strong coffee\\}
  \end{xlist}
\end{exe}

@@

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collostructional Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least in part from the relative frequency with which the same lemmas are used in the non-alternating constructions.
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be used in the corpus study Section~\ref{sec:corpusstudies}.

@@

Third, as described above, the grammaticalisation path leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- inasmuch as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified.
However, it is much more prototypical of quantities to be specified with numerical precision.
This is most obvious with the highly grammaticalised physical measure nouns like \textit{centilitre} which are very prototypically used with exact numeral instead of unspecific quantifiers, although both options are available (\textit{three centilitres} vs.\ \textit{several centilitres}).
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected alternants under this hypothesis.

@@

Fourth and finally, it was found that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising, given that the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using two very simple proxy variables.

@@

% In \citet[10]{GelmanShalizi2013}, the authors -- prominent Bayesians themselves -- acknowledge that a theory of statistical testing is a desideratum, and they state about the standard inductive interpretation of Bayesianism that ``most of this received view of Bayesian inference is wrong'', and they develop a Bayesian notion of p-values (see also \citealp{Mayo2013}, for a frequentist reply; also \citealp{Senn2011} on different strands of Bayesianism and their stance on inductive vs.\ deductive reasoning, and \citealp{Mayo2011}, for a critical reply to \citealp{Senn2011}).
